\Large\textbf{Advection versus diffusion in brain ventricular transport}

### 2025-04-10 paper/main.tex
update cylinder convergence rates using hmin

Assuming that viscous forces are dominating the ventricular CSF flow,
we model the flow by the Stokes equations.
The net forces of motile cilia acting on the CSF are modeled as a steady,
bidirectional tangential traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}), with an impermeability condition at the walls.
Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($26.7$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, f}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{g}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $4.8$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

@@

In the computational model, we represent the \emph{smh} mutants by
considering transport of Dendra2 with CSF flow driven by cardiac
pressure pulsations only, representing the absence of motile
cilia. Comparing the in silico mutants with transport predictions in
the presence of both cardiac pulsations and motile cilia, we observe
noticeably delayed dynamics in ROIs 2, 4 and 5, but little change for
ROIs 3 and 6 (\Crefpanels{fig:fig4}{c}). Compared to experimental
data, the control and mutant scenarios differed more for the
simulations, with the times-to-threshold in ROI 2 and 4 being 81\% and
28\% higher in the absence of cilia-mediated flow. Moreover, the
predicted mean concentrations in ROI 5 deviate between the two
scenarios, which is not the case in the experiments. Altogether, our
findings identify that, for smaller proteins such as Dendra2,
diffusion plays an important role for solute movement to regions
upstream in the cilia-mediated flow fields or to distant ventricles,
while the cilia affect the downstream mean concentrations.
\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$
    in each region of interest (ROI) for control (black lines)
    and \emph{smh} mutant (red lines) zebrafish. 
    Bold lines are mean values of the cohorts and shaded regions cover one standard deviation.
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for controls (gray dots) and
    \emph{smh} mutants (red dots).
    The thick black lines denote mean values. The $p$-values
    were calculated using a Mann-Whitney U test with 95\% confidence level,
    under the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-induced/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in panels \textbf{a} and \textbf{c} the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

@@

To obtain $\Delta F$ values for each region of interest (ROI), ROIs were
first drawn on the aligned and block-averaged time series (\Cref{fig:fig2}).
Six ROIs were drawn manually: ROI 1 around the location of photoconversion,
and ROIs 2--6 in different regions of the ventricular system.
The pixel values of the fluorescence intensity within one ROI were
averaged for each time point, and the relative change $\Delta F$ calculated with this averaged value.
To normalize the data with respect to photoconversion efficiency, the
$\Delta F$ curves for all ROIs were divided by the $\Delta F$ values
obtained for ROI 1 at the end of the experiment,
so that $\Delta F$ values for the photoconverted site ROI 1 ranged from
zero to one. Finally, we averaged the $\Delta F$ data from controls and from mutants.

@@

To computationally represent protein photoconversion in a region
$\Omega_{\mathrm{pc}} \subset \Omega$ (corresponding to ROI 1),
we prescribe a given time-dependent value to the solute concentration in this region. 
Mimicking the experimental protocol, we consider the dorsal part of the
diencephalic ventricle as the subdomain $\Omega_{\mathrm{pc}}$, and set
\begin{equation}
  c(\xx, t)=\frac{\log{(1+t/a)}}{\log{(1+T/a)}},
  \quad \xx \in \Omega_{\mathrm{pc}}, \quad t > 0.
\label{eq:photoconversion_curve}
\end{equation}
The value $a=65$ was chosen to fit the fluorescence intensity curve
observed in the physical photoconversion experiments. Because steep gradients
in the concentration field may arise from imposing $c(\xx, t)$ in $\Omega_{\mathrm{pc}}$,
the mesh is locally refined around this region.

In order to impose a smooth initial condition, we initially solve the diffusion equation
\begin{equation}\label{eq:cinit}
  \begin{aligned}
    &\pdifft{c} = \Tilde{D}\nabla^2 c \quad \text{in } \Omega\times(0, 2\Delta t]\\
      &c(\xx, t=0) = 0\quad\text{ in }\Omega
  \end{aligned}\
\end{equation}
with an increased diffusion coefficient $\Tilde{D}=D\times 10^{5}$ for two timesteps.
The solutions are normalized by their respective maximum values, and subsequently
multiplied with the value of \Cref{eq:photoconversion_curve} at the times $t=\Delta t$
and $t=2\Delta t$, respectively.
The resulting concentration fields are used as initial conditions for $c(\xx, 0)$ and
$c(\xx, \Delta t)$ for all $\xx\in\Omega$, where $\Delta t$
is the timestep. The two initial conditions are required because of the
two-step temporal discretization scheme employed (see~\Cref{subsec:num_approx_adv_diff} for more details.)

@@

\subsection{Numerical verification of model implementation}\label{subsec:numerical_verification}
A cylinder geometry meshed with varying degrees of refinement was used to
perform convergence analysis of the numerical methods used to solve the Stokes equations.
The cylinder had unit length and a diameter of one half length unit.
For the analysis, we used tangential traction boundary conditions on all
of the cylinder surface. Using the Method of Manufactured Solutions~\cite{Roache2001CodeSolutions}
with the velocity and pressure expressions defined in~\Cref{subsec:mms}
for $\uu_{\mathrm{ex}}$ and $p_{\mathrm{ex}}$, we calculated the error of 
the velocity approximation $\uu_h$
in the $\mathbf{L}^2$ norm and $\mathbf{H}^1$ semi-norm,
and the error of the pressure approximation $p_h$ in the $L^2$ norm (\Cref{tab:cylinder_error_rates}).
Based on the error calculations,
we calculated the experimental order of convergence,
as defined in \Cref{subsec:norms_and_eoc}.
We observe convergence in all of the reported error norms. Furthermore,
we report the error in calculation of the maximum velocities, and observe
convergence in this error as well. Finally, we calculated the $L^2$ norm of
the divergence $\nabla\cdot\uu_h$ to verify the property of mass conservation.
\begin{table}[!htbp]
    \footnotesize
    \centering
    \caption{Error norms for the convergence study
    performed with a cylinder mesh with
    the cilia-driven/no-cardiac flow model using $\mu=1$.
    For all error norms, convergence rates calculated
    with~\Cref{eq:convergence_order_estimate} are given
    in parentheses. The coarsest mesh had a mesh size $h_0$,
    and for every mesh refinement the mesh size $h$ was halved.}\label{tab:cylinder_error_rates}
    \begin{tabular}{cc|ccccc}
        \toprule
        $h/h_0$ & \# dofs &
        $\normltwovec{\uu_h - \uu_{\mathrm{ex}}}$
        & $\Hnorm{\uu_h - \uu_{\mathrm{ex}}}$
        & $\normltwo{p_h - p_{\mathrm{ex}}}$
        & $\vert\normlinf{u_h} - \normlinf{u_{\mathrm{ex}}}\vert$
        & $\normltwo{\nabla\cdot\uu_h}$\\ 
        \midrule 
        1    & 8652    & \num{2.3e-02}  (--)  & \num{4.3e-01}   (--)  & \num{5.6e-01}  (--)  & \num{2.4e-02}  (--)  & \num{1.4e-07} \\  
        1/2  & 41190   & \num{9.3e-03} (1.30) & \num{2.8e-01}  (0.61) & \num{3.9e-01} (0.53) & \num{5.0e-03} (2.25) & \num{2.8e-08} \\  
        1/4  & 255016  & \num{3.5e-03} (1.42) & \num{1.4e-01}  (1.00) & \num{2.1e-01} (0.85) & \num{1.4e-03} (1.83) & \num{7.3e-11} \\     
        1/8  & 1855337 & \num{1.3e-03} (1.46) & \num{6.9e-02}  (1.03) & \num{1.1e-01} (0.92) & \num{6.7e-04} (1.08) & \num{1.5e-10} \\   
        \bottomrule
    \end{tabular}
\end{table}

@@

The $\mathrm{BDM}_1$-$\mathrm{DG}_0$ scheme that we employ for the Stokes equations
is computationally more expensive than a more conventional approach based on
e.g. Taylor-Hood discretization~\cite{Stenberg1990ErrorProblem}
($P_2-P_1$ elements) or lower-order enriched
elements~\cite{Brezzi2011MixedMethods},
such as the MINI-element.
We opted for the BDM-DG scheme because of its exactly-mass-conserving property,
since numerical experiments proved that a divergence-free velocity field was
fundamental to attain a stable discretization of the advection-diffusion equation
in advection-dominated regions. First, we attempted
to discretize the Stokes equations with 
Taylor-Hood elements, but with the resulting numerical approximation
of the velocity $\uu_h$, we were unable to achieve
stability of the advection-diffusion discretization.
Numerical experiments identified non-zero divergence of $\uu_h$
as the root cause of these stability issues.
On the standard computational mesh, the $L^2$ norm of $\nabla\cdot\uu_h$ was
in the order of $10^{-6}$, while the
maximum value of the divergence of the velocity was of order unity.
On a refined mesh with roughly six times as many
computational cells as the standard mesh, the numerical values of
$\normltwo{\nabla\cdot\uu_h}$ and $\normlinf{\nabla\cdot\uu_h}$
remained practically unchanged. This is a result
of the fact that incompressibility is only achieved in a weak sense when
employing Taylor-Hood elements. The advective part of the total flux
is
\begin{equation*}
    \nabla\cdot(c\uu) = c(\nabla\cdot\uu) + \uu\cdot\nabla c.
\end{equation*}
In the case that $\nabla\cdot\uu\neq 0$, a source (or sink) term for the concentration
is introduced. This may introduce errors in the numerical approximation of $c$
and lead to instabilities and spurious oscillations, especially in the case of
advection-dominated transport
\cite{gresho1998incompressible, cesmelioglu2022compatible, johnson2009numerical}.
The $\mathrm{BDM}_1$-$\mathrm{DG}_0$ scheme avoids these issues,
since $\nabla\cdot\uu_h=0$ is satisfied exactly as a result of the element definitions.

>>>>

Assuming that viscous forces are dominating the ventricular CSF flow,
we model the flow by the Stokes equations.
The net forces of motile cilia acting on the CSF are modeled as a steady,
bidirectional tangential traction acting on parts of the ventricular wall
(\Crefpanels{fig:fig1}{d}), with an impermeability condition (no normal flow) at the walls.
Simulations show that this traction alone
leads to a partial compartmentalization of the ventricular system,
with large-scale vortex structures in the anterior, middle and
posterior ventricles (\Crefpanels{fig:fig1}{e}). The highest flow
speeds ($26.7$ \textmu m/s) occur in the vicinity of the dorsal cilia
in the middle ventricle. To model the cardiac pulsations, we impose a
pulsatile pressure difference between the anterior and posterior
boundaries (\Crefpanels{fig:fig1}{d, f}). The resulting flow
is pulsatile and directional (\Crefpanels{fig:fig1}{g}), and
in the absence of the cilia, the vortex structures vanish. The
velocity magnitude peaks at $4.8$ \textmu m/s. CSF initially flows
rostrocaudally before changing direction to caudorostral in the middle
of the cardiac cycle.

@@

In the computational model, we represent the \emph{smh} mutants by
considering transport of Dendra2 with CSF flow driven by cardiac
pressure pulsations only, representing the absence of motile
cilia. Comparing the in silico mutants with transport predictions in
the presence of both cardiac pulsations and motile cilia, we observe
noticeably delayed dynamics in ROIs 2, 4 and 5, but little change for
ROIs 3 and 6 (\Crefpanels{fig:fig4}{c}). Compared to experimental
data, the control and mutant scenarios differed more for the
simulations, with the times-to-threshold in ROI 2 and 4 being 81\% and
28\% higher in the absence of cilia-mediated flow. Moreover, the
predicted mean concentrations in ROI 5 deviate between the two
scenarios, which is not the case in the experiments. Altogether, our
findings identify that, for smaller proteins such as Dendra2,
diffusion plays an important role for solute movement to regions
upstream in the cilia-mediated flow fields or to distant ventricles,
while the cilia affect the downstream mean concentrations.
\begin{figure}[H]
    \centering
    \includegraphics[width=0.96\textwidth]{graphics/figure4_compare_control_mutant.png}
    \caption{\textbf{a.} Relative change in fluorescence intensity $\Delta F(t)$
    in each region of interest (ROI) for control (black lines)
    and \emph{smh} mutant (red lines) zebrafish. 
    Bold lines are mean values of the cohorts and shaded regions cover one standard deviation.
    \textbf{b.} Times when $\Delta F(t)$ first exceeds a threshold value,
    equal to 0.25 (ROIs 2--4) or 0.10 (ROIs 5, 6), for controls (gray dots) and
    \emph{smh} mutants (red dots).
    The thick black lines denote mean values. The $p$-values
    were calculated using a Mann-Whitney U test with 95\% confidence level,
    under the null hypothesis that the two distributions of control and mutant data are 
    from the same population.
    \textbf{c.} Mean concentration $\cbar(t)$ in each ROI,
    simulated using the baseline model (black lines) and the cardiac-induced/no-cilia model (red lines).
    The horizontal gray lines mark the threshold value $\hat{c}$, and the time
    $\hat{t}$ when the threshold is first exceeded is reported in the plot legends.
    Note that in panels \textbf{a} and \textbf{c} the vertical axes have
    different limits in each plot.}
    \label{fig:fig4}
\end{figure}

@@

To obtain $\Delta F$ values for each region of interest (ROI), ROIs were
first drawn on the aligned and block-averaged time series (\Cref{fig:fig2}).
Six ROIs were drawn manually: ROI 1 around the location of photoconversion,
and ROIs 2--6 in different regions of the ventricular system.
The pixel values of the fluorescence intensity within one ROI were
averaged for each time point, and the relative change $\Delta F$ calculated with this averaged value.
To normalize the data with respect to photoconversion efficiency, the
$\Delta F$ values for all ROIs were divided by the $\Delta F$ values
obtained for ROI 1 at the end of the experiment,
so that $\Delta F$ values for the photoconverted site ROI 1 ranged from
zero to one. Finally, we averaged the $\Delta F$ data from controls and from mutants.

@@

To computationally represent protein photoconversion in a region
$\Omega_{\mathrm{pc}} \subset \Omega$ (corresponding to ROI 1),
we prescribe a given time-dependent value to the solute concentration in this region. 
Mimicking the experimental protocol, we consider the dorsal part of the
diencephalic ventricle as the subdomain $\Omega_{\mathrm{pc}}$, and set
\begin{equation}
  c(\xx, t)=\frac{\log{(1+t/a)}}{\log{(1+T/a)}},
  \quad \xx \in \Omega_{\mathrm{pc}}, \quad t\in(0, T].
\label{eq:photoconversion_curve}
\end{equation}
The value $a=65$ was chosen to fit the fluorescence intensity curve
observed in the physical photoconversion experiments. Because steep gradients
in the concentration field may arise from imposing $c(\xx, t)$ in $\Omega_{\mathrm{pc}}$,
the mesh is locally refined around this region.

In order to impose a smooth initial condition, we initially solve the diffusion equation
\begin{alignat*}{2}
\pdifft{c} &= \Tilde{D}\nabla^2 c \quad &&\text{in } \Omega\times(0, 2\Delta t]\\
    c(\xx, t=0) &= 0 \quad&&\text{in }\Omega
\end{alignat*}
with an increased diffusion coefficient $\Tilde{D}=D\times 10^{5}$ for two timesteps.
The solutions are normalized by their respective maximum values, and subsequently
multiplied with the value of \Cref{eq:photoconversion_curve} at the times $t=\Delta t$
and $t=2\Delta t$, respectively.
The resulting concentration fields are used as initial conditions for $c(\xx, 0)$ and
$c(\xx, \Delta t)$ for all $\xx\in\Omega$, where $\Delta t$
is the timestep. The two initial conditions are required because of the
two-step temporal discretization scheme employed (see~\Cref{subsec:num_approx_adv_diff} for more details.)

@@

\subsection{Numerical verification of model implementation}\label{subsec:numerical_verification}
A cylinder geometry meshed with varying degrees of refinement was used to
perform convergence analysis of the numerical method used to solve the Stokes equations.
The cylinder had unit length and a diameter of one half length unit.
For each mesh refinement, the edge lengths were approximately halved.
Since the refinement was not uniform, we used the ratio of the minimal
edge lengths $h_{\mathrm{min}, k}/h_{\mathrm{min}, k+1}$ between two 
successive refinements when calculating the convergence rate using
\Cref{eq:convergence_order_estimate}.
For the analysis, we used tangential traction boundary conditions on all
of the cylinder surface. Using the Method of Manufactured Solutions~\cite{Roache2001CodeSolutions}
with the velocity and pressure expressions defined in~\Cref{subsec:mms}
for $\uu_{\mathrm{ex}}$ and $p_{\mathrm{ex}}$, we calculated the error of 
the velocity approximation $\uu_h$
in the $\mathbf{L}^2$ norm and $\mathbf{H}^1$ semi-norm,
and the error of the pressure approximation $p_h$ in the $L^2$ norm (\Cref{tab:cylinder_error_rates}).
Based on the error calculations,
we calculated the experimental order of convergence,
as defined in \Cref{subsec:norms_and_eoc}.
We observe convergence in all of the reported error norms. Furthermore,
we report the error in calculation of the maximum velocities, and observe
convergence in this error as well. Finally, we calculated the $L^2$ norm of
the divergence $\nabla\cdot\uu_h$ to verify the property of mass conservation.
\begin{table}[!htbp]
    \footnotesize
    \centering
    \caption{Error norms for the convergence study
    performed with a cylinder mesh with
    the cilia-driven/no-cardiac flow model using $\mu=1$.
    For all error norms, convergence rates calculated
    with~\Cref{eq:convergence_order_estimate} are given
    in parentheses.}\label{tab:cylinder_error_rates}
    \begin{tabular}{cc|ccccc}
        \toprule
        \# cells & \# dofs &
        $\normltwovec{\uu_h - \uu_{\mathrm{ex}}}$
        & $\Hnorm{\uu_h - \uu_{\mathrm{ex}}}$
        & $\normltwo{p_h - p_{\mathrm{ex}}}$
        & $\vert\normlinf{u_h} - \normlinf{u_{\mathrm{ex}}}\vert$
        & $\normltwo{\nabla\cdot\uu_h}$\\ 
        \midrule 
        1140    & 8652    & \num{2.3e-02}  (--)  & \num{4.3e-01}   (--)  & \num{5.6e-01}  (--)  & \num{2.4e-02}  (--)  & \num{1.4e-07} \\  
        5568    & 41190   & \num{9.3e-03} (1.59) & \num{2.8e-01}  (0.75) & \num{3.9e-01} (0.65) & \num{5.0e-03} (2.75) & \num{2.8e-08} \\  
        35311   & 255016  & \num{3.5e-03} (1.58) & \num{1.4e-01}  (1.11) & \num{2.1e-01} (0.95) & \num{1.4e-03} (2.03) & \num{7.3e-11} \\     
        260717  & 1855337 & \num{1.3e-03} (1.59) & \num{6.9e-02}  (1.12) & \num{1.1e-01} (1.00) & \num{6.7e-04} (1.18) & \num{1.5e-10} \\   
        \bottomrule
    \end{tabular}
\end{table}

@@

The $\mathrm{BDM}_1$-$\mathrm{DG}_0$ scheme that we employ for the Stokes equations
is computationally more expensive than more common approaches based on
e.g. Taylor-Hood discretization~\cite{Stenberg1990ErrorProblem}
($P_2-P_1$ elements) or lower-order enriched
elements (e.g. MINI)~\cite{Brezzi2011MixedMethods}.
We opted for the BDM-DG scheme because of its exactly-mass-conserving property,
since numerical experiments proved that a divergence-free velocity field was
fundamental to attain a stable discretization of the advection-diffusion equation
in advection-dominated regions. First, we attempted
to discretize the Stokes equations with 
Taylor-Hood elements, but with the resulting numerical approximation
of the velocity $\uu_h$, we were unable to achieve
stability of the advection-diffusion discretization.
Numerical experiments identified non-zero divergence of $\uu_h$
as the root cause of these stability issues.
On the standard computational mesh, the $L^2$ norm of $\nabla\cdot\uu_h$ was
in the order of $10^{-6}$, while the
maximum value of the divergence of the velocity was of order unity.
On a refined mesh with roughly six times as many
computational cells as the standard mesh, the numerical values of
$\normltwo{\nabla\cdot\uu_h}$ and $\normlinf{\nabla\cdot\uu_h}$
remained practically unchanged. This is a result
of the fact that incompressibility is only achieved in a weak sense when
employing Taylor-Hood elements. The advective part of the total flux
is
\begin{equation*}
    \nabla\cdot(c\uu) = c(\nabla\cdot\uu) + \uu\cdot\nabla c.
\end{equation*}
In the case that $\nabla\cdot\uu\neq 0$, a source (or sink) term for the concentration
is introduced. This may introduce errors in the numerical approximation of $c$
and lead to instabilities and spurious oscillations, especially in the case of
advection-dominated transport
\cite{gresho1998incompressible, cesmelioglu2022compatible, johnson2009numerical}.
The $\mathrm{BDM}_1$-$\mathrm{DG}_0$ scheme avoids these issues,
since $\nabla\cdot\uu_h=0$ is satisfied exactly as a result of the element definitions.
